Generalization Errors of Hard Parameter Sharing: High-dimensional Asymptotics and Implications

### 2020-09-24 arxiv/implications.tex
revise

%We provide precise explanations to the phenomena of negative transfer in multi-task learning.
%We provide tight bounds on the bias and variance of the multi-task estimator for two tasks.
%The results in Section \ref{sec_general} show when multi-task learning outperforms single-task learning for general task covariance matrices and ground truth model parameters.
%In particular, the general conditions depend on specific properties of task data such as the covariate shift matrix.
%However, the conditions given by the limiting bias and variance of multi-task learning can be difficult to interpret.
%The goal of this section is to address this issue.
%We interpret the limiting bias and variance equations in an isotropic covariance setting that is described below and a simplified covariate-shifted setting.
%We show that three properties of task data affect the performance of multi-task learning as follows.
%(i) \textit{Task similarity}: We explain the phenomenon of negative transfer precisely as tasks become more different.
%(ii) \textit{Sample ratio}: We further explain a curious phenomenon where increasing the sample ratio between the source and target task helps initially, but hurts eventually.
%(iii) \textit{Covariate shift}: Finally, we show that as the sample size of the source task increases, we show that the covariate shift worsens the performance of the multi-task estimator.
%Finally, we provide a extend our results from two tasks to many tasks with the same features.
%We explain from three perspectives, including \textit{task similarity}, \textit{sample size} and \textit{covariate shift}.
%We show how negative transfer occurs by varying task similarity or sample size.
%Then we show that when source task sample size becomes large, covariate shift causes more negative effects.
We illustrate our main results (to be described in Section \ref{sec_general}) in a simplified setting.
For this setting, we simulate several empirical phenomena that are well-known in hard parameter sharing approaches.
We use our results to explain the empirical phenomena.

@@

Using our techniques described later in Section \ref{sec_general}, we derive the limiting prediction loss of the hard parameter sharing estimator.
Without loss of generality, we describe the result for task two and remark that a similar result for task one holds.

\begin{corollary}[Derived from Lemma \ref{lem_cov_shift} and Lemma \ref{lem_cov_derivative}]
	In the isotropic setting described above, with high probability, the prediction loss of the hard parameter sharing estimator for task two satisfies:
	\begin{align*}
		& \bigabs{ L(\hat{\beta}_2^{\MTL}) - \frac{\sigma^2}{\rho_1 + \rho_2 - 1} - (1 + \frac{6}{\rho_1} + \frac{1}{\rho_1^2})pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} } \\
		\le & 	\frac{4}{\sqrt{\rho_1}}(1 + \frac{1}{\rho_1})	pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} + \oo(\sigma^2 + pd^2).
%	 \frac{\sigma^2}{\rho_1+\rho_2-1}+\left( 1-\frac{1}{\sqrt{\rho_1}}\right)^4  +\oo(\sigma^2 + pd^2)  \le \\
%	 \le  \frac{\sigma^2}{\rho_1+\rho_2-1}+\left( 1+\frac{1}{\sqrt{\rho_1}}\right)^4 pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} +\oo(\sigma^2 + pd^2)
	 \end{align*}
\end{corollary}

It is well-known since the seminal work of Caruana \cite{C97} that how well multi-task learning performs depends on task relatedness.
We formalize this connection in the following simplified setting, where we can perform explicit calculations.
We show that as we increase the distance between $\beta_1$ and $\beta_2$, there is a transition from positive transfer to negative transfer in MTL.
In particular, the bias of MTL increases as the distance between $\beta_1$ and $\beta_2$ increases.
Therefore, while the variance of MTL still reduces compared to STL, if the bias increases more than the amount of variance reduced, we will observe negative transfer.

>>>>

%We provide precise explanations to the phenomena of negative transfer in multi-task learning.
%We provide tight bounds on the bias and variance of the multi-task estimator for two tasks.
%The results in Section \ref{sec_general} show when multi-task learning outperforms single-task learning for general task covariance matrices and ground truth model parameters.
%In particular, the general conditions depend on specific properties of task data such as the covariate shift matrix.
%However, the conditions given by the limiting bias and variance of multi-task learning can be difficult to interpret.
%The goal of this section is to address this issue.
%We interpret the limiting bias and variance equations in an isotropic covariance setting that is described below and a simplified covariate-shifted setting.
%We show that three properties of task data affect the performance of multi-task learning as follows.
%(i) \textit{Task similarity}: We explain the phenomenon of negative transfer precisely as tasks become more different.
%(ii) \textit{Sample ratio}: We further explain a curious phenomenon where increasing the sample ratio between the source and target task helps initially, but hurts eventually.
%(iii) \textit{Covariate shift}: Finally, we show that as the sample size of the source task increases, we show that the covariate shift worsens the performance of the multi-task estimator.
%Finally, we provide a extend our results from two tasks to many tasks with the same features.
%We explain from three perspectives, including \textit{task similarity}, \textit{sample size} and \textit{covariate shift}.
%We show how negative transfer occurs by varying task similarity or sample size.
%Then we show that when source task sample size becomes large, covariate shift causes more negative effects.
We illustrate our main results, Theorem \ref{thm_many_tasks}, Theorem \ref{thm main RMT} and Corollary \ref{thm_main_informal}, in a simplified setting.
For this setting, we simulate several empirical phenomena that are well-known in hard parameter sharing approaches.
We use our results to explain the empirical phenomena.

@@

Using Theorem \ref{thm main RMT}, %our techniques described later in Section \ref{sec_general}, 
we derive the limiting prediction loss of the hard parameter sharing estimator.
Without loss of generality, we describe the result for task two and remark that a similar result for task one holds.

\begin{corollary}[Derived from Theorem \ref{thm main RMT}]
	Assume that the isotropic setting described above holds, and that there exists a small constant $c_0>0$ such that the signal-to-noise ratios satisfy that
\be\label{choiceofpara0}
p^{ c_0} \le \frac{p\kappa^2}{\sigma^2 }  \le p^{1-c_0},\quad  \frac{pd^2}{\sigma^2}=\OO(1).
\ee
	Let $B, W_1, W_2$ be the global minimum of equation \eqref{eq_mtl}. Then with high probability, the prediction loss of the hard parameter sharing estimator for task two satisfies:
	\begin{align*}
	 \left( 1-\frac{1}{\sqrt{\rho_1}}\right)^4 pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} +\OO(p^{-c}\sigma^2)  \le L(\hat{\beta}_2^{\MTL}) -\frac{\sigma^2}{\rho_1+\rho_2-1} \\
	 \le  \left( 1+\frac{1}{\sqrt{\rho_1}}\right)^4 pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} +\OO(p^{-c}\sigma^2 )
	 \end{align*}
	 for any constant $c\in (0,c_0)$.
	 \iffalse
	\begin{align*}
		& \bigabs{ L(\hat{\beta}_2^{\MTL}) - \frac{\sigma^2}{\rho_1 + \rho_2 - 1} - (1 + \frac{6}{\rho_1} + \frac{1}{\rho_1^2})pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} } \\
		\le & 	\frac{4}{\sqrt{\rho_1}}(1 + \frac{1}{\rho_1})	pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} + \oo(\sigma^2 + pd^2).
%	 \frac{\sigma^2}{\rho_1+\rho_2-1}+\left( 1-\frac{1}{\sqrt{\rho_1}}\right)^4  +\oo(\sigma^2 + pd^2)  \le \\
%	 \le  \frac{\sigma^2}{\rho_1+\rho_2-1}+\left( 1+\frac{1}{\sqrt{\rho_1}}\right)^4 pd^2\cdot \frac{\rho_1^2 (\rho_1+\rho_2)}{(\rho_1 + \rho_2 - 1)^3} +\oo(\sigma^2 + pd^2)
	 \end{align*}
	 \fi
\end{corollary}

It is well-known since the seminal work of Caruana \cite{C97} that how well multi-task learning performs depends on task relatedness.
We formalize this connection in the above isotropic setting, where we can perform explicit calculations.
We show that as we increase the distance between $\beta_1$ and $\beta_2$, there is a transition from positive transfer to negative transfer in MTL.
In particular, the bias of MTL increases as the distance between $\beta_1$ and $\beta_2$ increases.
Therefore, while the variance of MTL still reduces compared to STL, if the bias increases more than the amount of variance reduced, we will observe negative transfer.
